--- 
title: Homework 1
subtitle: p-values
author: Warren Kennedy
date: today
toc: true
---

```{r}
#| message:false
#| echo:false

library(tidyverse)
```


# Section 1: Basic text formatting

## Sub-section

### Header

- some
- words
- in
- bullets


1. some
2. words
3. in
4. numbers

text in *italics* and **bold**, or ~~strikethrough~~

hyperlinks: <https://www.google.com> or [click here](https://www.google.com)

# Section 2: some math

in line math: $y = mx + b$

$$
\int_0^1 x^2 dx = \frac{1}{3}
$$


some latex code:

```{=latex}
\begin(align)
y &= mx + b \\ 
 &=ab + ac 
\end{align}
```

# Section 3: code

you can do inline code like `r format(sqrt(2), digits=3)` or code blocks:

```{r}
x <- 5
y <- 10
z <- x + y
```


```{r}
#| results=hide

data(mtcars)
cor(mtcars$mpg, mtcars$wt)
```



```{r}
#| echo: false

data(mtcars)
cor(mtcars$mpg, mtcars$wt)
```



```{r}
#| eval: false

data(mtcars)
cor(mtcars$mpg, mtcars$wt)
```


# Section 4: plots


```{r}
#| fig-height: 6
#| fig-width: 6
#| fig-align: center
mtcars %>% 
  ggplot(aes(x=wt, y=mpg)) +
  geom_point(col="dodgerblue", size=3)
```


# Section 5: tables

```{r}
mtcars[1:5,]
```


```{r}
knitr::kable(mtcars[1:5,],)
```


```{r}
gt::gt(mtcars[1:5,],)
```


```{r}
stargazer::stargazer(mtcars[1:5,],)
```


# Bonus Section

:::{.callout-note collapse="true"}
## note title

note body




## Bonus 2: Panels

:::



a tabset panel

:::{.panel-tabset}

## Jesus

```{r}
p + geom_smooth(method="lm", se=F)
```

# Brian 

```{r}
p + geom_smooth(method="loess", se=F)
```

:::